---
title: "Divvy Bike Analysis"
format: 
  html:
    theme: litera
    code-fold: true
    code-summary: "Show Code"
    code-overflow: wrap
    code-tools: true
    code-copy: true
    highlight-style: github
    code-link: true
    
---


## Introduction

Public bike share (PBS).

A PBS is an extension of the existing transit system with a network of
short-term, self-service bicycle posts in which:

-   Users can use bikes by purchasing casual day use or annual
    memberships,

-   Users can ride bikes for a short distance for one-way trips within a
    defined service area and Station locations can change overtime based
    on ridership practices or operation needs.

Bike share technology continues to evolve quickly along with other
wireless and digital changes. Other recent advancements include systems
that do not require docking stations i.e. "smart lock" systems and
electric‐assist bikes.

### About this dataset

I used 1 year (12 Months) of Divvy Chicago Bike Sharing Data from Jan
2024 to Dec 2024, obtained from Divy Bike .

This dataset included 5,667,986 trips, of which 3,127,293 / 55.2% were
made by annual or monthly members, and 2,540,693 / 44.8% by casual
users.

#### Missing features

Although this dataset is not a result of a survey, The following missing
features could make this analysis more interesting and stronger if were
exist.

-   Bike ID,
-   User ID,
-   User Gender,
-   User Age,
-   Purpose of trip,
-   Trip distance and
-   Weather conditions.

I believe some of them are connected to the agreed privacy policy!

Anyway, let's load the packages we need for this brief analysis

```{r setup, class.source="Fragment Mono", results='hide', warning=FALSE, message=FALSE}
#| code-fold: true
#| code-summary: "Show the code"
#| fold: true

knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(plotly)
library(ggplot2)
library(paletteer)
library(PNWColors) 
library(gghighlight)
library(networkD3)
scale_colour_paletteer_d("PNWColors::Sunset")
scale_fill_paletteer_d("PNWColors::Sunset")
theme_set(theme_minimal())
```
We are setting up our system by loading the necessary libraries. These libraries will enable us to perform data manipulation, visualization, and interactive chart creation.
## Reading Data
In this section, we are reading the Divvy bike data for all months of 2024. Each dataset represents trips taken in a specific month. The data is stored in CSV files located in the `datasets` folder.

```{r reading all the datasets, class.source="Fragment Mono", results='hide', warning=FALSE, message=FALSE}
#| code-fold: true
#| code-summary: "Show the code"
#| fold: true
january <- read.csv("datasets/202401-divvy-tripdata.csv")
february <- read.csv("datasets/202402-divvy-tripdata.csv")
march <- read.csv("datasets/202403-divvy-tripdata.csv")
april <- read.csv("datasets/202404-divvy-tripdata.csv")
may <- read.csv("datasets/202405-divvy-tripdata.csv")
june <- read.csv("datasets/202406-divvy-tripdata.csv")
july <- read.csv("datasets/202407-divvy-tripdata.csv")
august <- read.csv("datasets/202408-divvy-tripdata.csv")
september <- read.csv("datasets/202409-divvy-tripdata.csv")
october <- read.csv("datasets/202410-divvy-tripdata.csv")
november <- read.csv("datasets/202411-divvy-tripdata.csv")
december <- read.csv("datasets/202412-divvy-tripdata.csv")
```

## Combining the datasets


Here, we combine all monthly datasets into a single dataset for easier analysis. Additionally, a `month` column is added to each dataset to identify the month of each ride.

```{r combining all the datasets, class.source="Fragment Mono", results='hide', warning=FALSE, message=FALSE}

#| code-fold: true
#| code-summary: "Show the code"
#| fold: true
all_data <- bind_rows(
    january |> mutate(month = 1),
    february |>  mutate(month = 2),
    march|>  mutate(month = 3),
    april|>  mutate(month = 4),
    may|>  mutate(month = 5),
    june|>  mutate(month = 6),
    july |> mutate(month = 7),
    august |> mutate(month = 8),
    september |> mutate(month = 9),
    october |> mutate(month = 10),
    november |> mutate(month = 11),
    december|>  mutate(month = 12)
)
```

## Sample Dataset
To make our analysis more efficient and focus on a manageable subset of data, we sample 100 records from the combined dataset.

```{r taking a sample, class.source="Fragment Mono", results='hide', warning=FALSE, message=FALSE}
#| code-fold: true
#| code-summary: "Show the code"
#| fold: true
all_data <- all_data |>
    sample_n(100)
```

## Combining Data
To enhance our analysis, we derive new columns:
- **Hour**: Extracts the hour from the start time.
- **Time of Day**: Categorizes hours into time periods (Night, Morning, Afternoon, Evening).
- **Season**: Categorizes months into seasons (Winter, Spring, Summer, Fall).

```{r adding daytime and season, class.source="Fragment Mono", results='hide', warning=FALSE, message=FALSE}
#| code-fold: true
#| code-summary: "Show the code"
#| fold: true

all_data <- all_data |>
    mutate(hour = hour(started_at)) |>
    mutate(time_of_day = 
    case_when (
      hour %in% 0:5 ~ "Night",
      hour %in% 6:11 ~ "Morning",
      hour %in% 12:17 ~ "Afternoon",
      hour %in% 18:23 ~ "Evening"
        )
    )

all_data <- all_data  |>
    mutate(season = 
    case_when (
        month %in% c(12, 1, 2) ~ "Winter",
        month %in% c(3, 4, 5) ~ "Spring",
        month %in% c(6, 7, 8) ~ "Summer",
        month %in% c(9, 10, 11) ~ "Fall"
    )
)
sample_data <- all_data  |>
    sample_n(100)
```

## Remove Data
We clean up our workspace by removing the individual monthly datasets, which are no longer needed after combining.

```{r removing memory space, class.source="Fragment Mono", results='hide', warning=FALSE, message=FALSE}
#| code-fold: true
#| code-summary: "Show the code"
#| fold: true
rm(january, february, march, april, may, june, july, august, september, october, november, december)
```

## Ride type by user
This section analyzes ride types based on user type (member vs. casual). The data is grouped and visualized in a pie chart to show the proportion of rides for each user type.
```{r, ride type by user}
#| code-fold: true
#| code-summary: "Show the code"
#| fold: true
ride_type_by_user <- sample_data |>
    group_by(member_casual) |>
    summarise(count = n()) |>
    mutate(
        proportion = count / sum(count),
        percentage = count / sum(count) * 100,
    )

ride_type_by_user |>
    mutate(
        start = lag(proportion, default = 0) * 2 * pi,
        end = cumsum(proportion) * 2 * pi,
    ) |>
    ggplot() +
    ggforce::geom_arc_bar(
        aes(
            x0 = 0, y0 = 0,
            r0 = 0.7, r = 1,
            start = start, 
            end = end,
            fill = member_casual 
        )
    ) +
    annotate(
      'text',  
      x = 0,
      y = 0,
      label = paste0(round(nrow(sample_data) / 1000000, 2), "M\nTotal\nRides"),
      size = 7,
      lineheight = 1
    ) +
    coord_equal(expand = FALSE, xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
    theme_void(base_size = 16) +
    labs(
        fill = element_blank()
    ) +
    theme(
      legend.position = "none",
        legend.margin = margin(t = 1, b = 0.5, unit = 'cm'), 
    ) +
    scale_fill_manual(values = pnw_palette("Sunset", n = 2))
```

## Ride Type By Rideable
In this section, we examine ride type based on bike types (classic, electric, docked) and user categories. The data is visualized as a stacked bar chart to highlight usage trends.

```{r, ride type by rideable}
#| code-fold: true
#| code-summary: "Show the code"
#| fold: true

ride_type_by_rideable <- sample_data |>
    group_by(rideable_type, member_casual) |>
    summarise(count = n(), .groups = 'drop')

ggplot(ride_type_by_rideable, aes(x = rideable_type, y = count, fill = member_casual)) +
    geom_bar(stat = "identity", position = "stack", width = 0.7) +
    labs(
        x = "Bike Type",
        y = "Count",
        fill = "User Type"
    ) +
    theme_void(base_size = 15) +
    theme(
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
        legend.position = "top"
    ) + scale_fill_manual(values = pnw_palette("Sunset", n = 2))

   
```




```{r, average ride length}
## Average Ride Length
#| code-fold: true
#| code-summary: "Show the code"
average_ride_length <- sample_data |>
    mutate(ride_length = as.numeric(difftime(ended_at, started_at, units = "mins"))) |>
    group_by(member_casual) |>
    summarise(average_ride_length = mean(ride_length, na.rm = TRUE))

ggplot(average_ride_length, aes(x = member_casual, y = average_ride_length, fill = member_casual)) +
    geom_bar(stat = "identity", width = 0.7) +
    geom_text(aes(label = round(average_ride_length, 1)), vjust = -0.5, size = 5) +
    labs(
        x = "User Type",
        y = "Average Ride Length (minutes)"
    ) +
  scale_fill_manual(values = pnw_palette("Sunset", n = 2))
```

## Ride Length by Weekday
```{r ride length by weekday, height = 500, width = 800}
#| code-fold: true
#| code-summary: "Show the code"
ride_length_by_weekday <- sample_data |>
    mutate(ride_length = as.numeric(difftime(ended_at, started_at, units = "mins")),
           day_of_week = weekdays(as.Date(started_at))) |>
    group_by(day_of_week) |>
    summarise(average_ride_length = mean(ride_length, na.rm = TRUE)) |> 
    mutate(day_of_week = fct_relevel(day_of_week, c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")))

ggplot(ride_length_by_weekday, aes(x = day_of_week, y = average_ride_length, fill = day_of_week)) +
    geom_bar(stat = "identity", width = 0.7) +
    geom_text(aes(label = round(average_ride_length, 1)), vjust = -0.5, size = 5) +
    labs(
        x = "Day of the Week",
        y = "Average Ride Length (minutes)"
    ) +
  scale_fill_manual(values = pnw_palette("Sunset", n = 7))
```

## Total Rides By Weekday
```{r, total rides by weekday}
total_rides_by_weekday <- sample_data |>
    mutate(day_of_week = weekdays(as.Date(started_at))) |>
    group_by(day_of_week) |>
    summarise(total_rides = n()) |>
    mutate(day_of_week = fct_relevel(day_of_week, c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")))

total_rides_by_weekday |>
    ggplot(aes(x = day_of_week, y = total_rides, fill = day_of_week)) +
    geom_bar(stat = "identity", width = 0.7) +
    geom_text(aes(label = total_rides), vjust = -0.5, size = 5) +
    labs(
        x = "Day of the Week",
        y = "Total Rides"
    ) +
  scale_fill_manual(values = pnw_palette("Sunset", n = 7))
```

## Total Rides by Hour
```{r, total rides by hour}
total_rides_by_hour <- sample_data |>
    mutate(hour = hour(started_at)) |>
    group_by(hour, time_of_day) |>
    summarise(total_rides = n(), .groups = 'drop')

total_rides_by_hour |>
    ggplot(aes(x = hour, y = total_rides, fill = time_of_day)) +
    geom_bar(stat = "identity", width = 0.7) +
    scale_x_continuous(breaks = 0:23) +
    scale_fill_brewer(palette = "Set1") +
    labs(
        x = "Hour of the Day",
        y = "Total Rides",
        fill = "Time of Day"
    ) +
    theme_minimal(base_size = 15) + scale_fill_manual(values = pnw_palette("Sunset", n = 4))
```


## Total Rides by Month
```{r, total rides by month}

total_rides_by_month <- sample_data |>
    mutate(month = month(started_at, label = TRUE)) |>
    group_by(month, season) |>
    summarise(total_rides = n(), .groups = 'drop')

total_rides_by_month |>
    ggplot(aes(x = month, y = total_rides, fill = season)) +
    geom_bar(stat = "identity", width = 0.7) +
    scale_fill_brewer(palette = "Set1") +
    labs(
        x = "Month",
        y = "Total Rides",
        fill = "Season"
    ) + scale_fill_manual(values = pnw_palette("Sunset", n = 4))
```


```{r average ride length by season}

average_ride_length_by_season <- sample_data |>
    mutate(ride_length = as.numeric(difftime(ended_at, started_at, units = "mins"))) |>
    group_by(season) |>
    summarise(average_ride_length = mean(ride_length, na.rm = TRUE), .groups = 'drop')

ggplot(average_ride_length_by_season, aes(x = season, y = average_ride_length, fill = season)) +
    geom_bar(stat = "identity", width = 0.7) +
    geom_text(aes(label = round(average_ride_length, 1)), vjust = -0.5, size = 5, color = "white") +
    labs(
        x = "Season",
        y = "Average Ride Length (minutes)"
    ) +
    scale_fill_manual(values = pnw_palette("Sunset", n = 4))
```

## Top 10 Starting Station
```{r top 10 starting station}
sample_data |>
  mutate(start_station_name = fct_lump(start_station_name, 10)) |> 
  count(start_station_id, start_station_name, name = "counts", sort = T) |> 
  filter(!is.na(start_station_name),
         !is.na(start_station_id),
         start_station_name != "Other") |>
  mutate(start_station_name = fct_reorder(start_station_name, counts)) |>
  top_n(n=5) |>
  ggplot(aes(x = start_station_name, y = counts, fill = start_station_name)) +
  geom_bar(stat = "identity") +
  coord_flip() + 
  scale_fill_manual(values = pnw_palette("Sunset", n = 13))
```

## Top 10 Ending Stations

```{r, top 10 ending station}
sample_data |>
    mutate(end_station_name = fct_lump(end_station_name, 10)) |> 
    count(end_station_id, end_station_name, name = "counts", sort = T) |> 
    filter(!is.na(end_station_name),
                 !is.na(end_station_id),
                 end_station_name != "Other") |>
    mutate(end_station_name = fct_reorder(end_station_name, counts)) |> 
    top_n(n = 5) |>
    ggplot(aes(x = end_station_name, y = counts, fill = end_station_name)) +
    geom_bar(stat = "identity") +
    coord_flip() + 
    scale_fill_manual(values = pnw_palette("Sunset", n = 13))

```

